Bottum-up drivers of phytoplankton community diversity in the Cariaco Basin, Venezuela

### 2024-11-19 agujournaltemplate.tex
Start working on Methods

\title{Drivers of Phytoplankton Community Variability in the Cariaco Basin, Venezuela}

@@

@@

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{The CARIACO Time Series}
  %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time-Series Program was established to investigate the link between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin is a large ($\sim$ 160 km long, 70 km wide) and deep ($\sim$ 1,400 m) basin, located on the north-eastern Venezuelan continental shelf, bound to the north by a sill at a mean depth of about 100 m. It is a relatively productive area ($\sim$ 400 $g C m^{-2} y^{-1}$), with primary production and vertical particulate organic matter fluxes that are higher than BATS and HOT locations, leading to the accumulation of anoxic bottom waters below 200 m depth.

    \subsubsection{Niskin Bottle Dataset}
        Samples were collected as part of the Cariaco Ocean Time-Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.67° W).

    \subsubsection{CTD Dataset}

    \subsubsection{Phytoplankton Microscopy}

    \subsubsection{ERA 5 climate data for wind}

    \subsubsection{AMO and MEI v.2 Climate indices}

        %% METHODS
    \subsubsection{Z-Scores calculation}

    \subsubsection{Clustering and NMDS}

        \subsubsection{Gradient Forest}

@@

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/BDEF_depth_PLOT.pdf}
\caption{Linear regression of diversity indices to log-transformed Chlorophyll Biomass, Primary Production, Resource Use Efficiency for nitrate ($RUE_N$) and for phosphate ($RUE_P$). $RUE$ is calculated as the ratio of Chlorophyll to the available nutrient ($NO_3$ and $PO_4$ respectively). All variables show a significant increasing linear trend with Genus Richness, but none for the Shannon Index or Pielou\'s Evenness Index.
\newline
\textit{I included this plot to show something about the Diversity to Ecosystem Function relationship at CARIACO. This can be left out, if it does not make sense to discuss in the context of this manuscript, or much further simplified, by removing the correlations to Shannon and Pielou, since none of them are significant. All correlations to Genus Richness are below p=0.05. A more advanced way to quantify these relationships would be mixed linear models, which I could do, if that makes sense in the context of the other results. For now, I wanted to keep it simple.}}
\end{figure}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% REFERENCES and BIBLIOGRAPHY
%
% \bibliography{<name of your .bib file>} don't specify the file extension
% don't specify bibliographystyle
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\title{Bottum-up drivers of phytoplankton community diversity in the Cariaco Basin, Venezuela}

@@

        % below is from 2019 REVIEW
        Hydrographic observations (Figure 2) show strong seasonality and interannual variability in the upper 400 m in the Cariaco Basin. This is controlled by seasonal and interannual changes in the wind and in the intensity of the geostrophic Caribbean Current (Muller-Karger et al. 1989). Primary upwelling was strongest between November and May. In general, during upwelling, sea surface temperatures ranged between approximately 23◦Cand25◦C. The secondary upwelling that occurs between June and August is shorter (∼5 weeks) and is approximately 1.5\°C warmer than the primary upwelling. It is driven by Ekman transport and an intensification of the curl of the wind close to the southern coast of the Caribbean (Rueda-Roa \& Muller-Karger 2013, Rueda-Roa et al. 2018). Upwelled waters move offshore in plumes that extended as far as 50 km to more than 250 km from the coast (Muller-Karger et al. 1989, 2010). From July to October, there is strong stratification in the upper 50 m of the water column after the wind weakens and the rain and rivers deliver less saline waters (Figure 2).
        Sea surface temperatures ranged from approximately 28◦C to 30◦C during the period of stratification (Figure 2). The mixed-layer depth varied from less than 10 m to 35 m during the 21-year time series (Figure 3). Satellite sea surface temperature and ocean color observations show that upwelling pulsates throughout the year at the scale of days to weeks, leading to marked variation in primary production (Figure 3). CARIACO rigorously characterized the annual seasonal variation in primary productivity and stratification. The length of the time series was key in capturing interannual variations that spanned decades. Satellite sea surface temperature and field data show that upwelling was stronger between 1996 and 2002. After 2002, trade wind intensity became weaker (Taylor et al. 2012).
        The field observations, particularly those in the upper 50 m, showed what appeared to be a return to stronger upwelling conditions starting in 2014 (Figure 2a).

@@

\section{Materials and Methods}
% Here is text on Materials and Methods.
%
\subsection{Study Site}
  %Talk about the Basin shortly, then about the time series
    The Cariaco Ocean Time-Series Program was established to investigate the link between ocean surface processes and the sinking flux of particulate carbon. The Cariaco Basin is a large ($\sim$ 160 km long, 70 km wide) and deep ($\sim$ 1,400 m) basin, located on the north-eastern Venezuelan continental shelf, bound to the north by a sill at a mean depth of about 100 m. It is a relatively productive area ($\sim$ 400 $g C m^{-2} y^{-1}$), with primary production and vertical particulate organic matter fluxes that are higher than BATS and HOT locations, leading to the accumulation of anoxic bottom waters below 200 m depth.

    \subsubsection{In-Situ Data}
    Samples were collected as part of the Cariaco Ocean Time-Series Program in monthly research vessel cruises to a station located in the eastern Cariaco Basin (10.50° N, 64.67° W). % https://imars.usf.edu/?page_id=1027 (10°30’N, 64°40’W) ???

    % from 2019 Review:
    The CARIACO field program consisted of a monthly core oceanographic cruise, two sediment trap deployment/recovery cruises per year, and two microbiogeochemical cruises targeting microbial processes across the oxic–anoxic interface per year. Field logistics and partial sample analysis were coordinated by the Fundación La Salle de Ciencias Naturales on Margarita Island, Venezuela. Details of the history of sampling procedures in CARIACO as well as sampling schemes, analytical methods, and data quality control and intercalibration procedures have been described by \citeA{mullerkarger_annual_2001, thunell_particulate_2007, astor_yrene_m_handbook_2013, astor_interannual_2013, astor_sintesis_2014}.

    % below paragraph is from Taylor et al. 2012
    % I should similarly reference the other more comprehensive data protocols like they do
    % but also explain the methodology quickly, like they did
    Sampling and processing protocols for hydrography, NPP, Chla, phaeo, nutrient concentrations, and sediment trap deployment and processing appear in Muller-Karger et al. (8). These data are available from www.imars. usf.edu/CAR, the National Oceanographic Data Center (NODC) (www.nodc. noaa.gov), and the Biological and Chemical Oceanography Data Management Office (BCO-DMO) (http://bco-dmo.org/data). From late 1995 to early 1998, nutrient data were provided by William Senior (Universidad de Oriente, Cumana, Venezuela) and by Dr. K. Fanning (University of South Florida, St. Petersburg, FL) from mid-1997 to the present. Samples from eight cruises were analyzed by both laboratories; PO43− and Si[OH]4 data agreed very well ($±$5\%), but NOx and NH4+ did not. Therefore, the entire PO43− and Si[OH]4 dataset and only the USF NOx and NH4+ dataset were used here.

    Phytoplankton enumeration was conducted on water samples (500 mL) collected by Niskin bottles at depths of 1, 7, 15, 25, 35, 55 m and fixed with 5\% sodium tetraborate-buffered formaldehyde (final concentration). Quantitative analyses were performed using settling chambers and an inverted microscope according the Utermöhl method (38).

    % The below is from Pinckney et al. 2015
    % They have more detail about Phytoplankton, either use this, or reference them
    Samples were collected as part of the Cariaco Ocean Time-Series Program at a station located in the eastern Cariaco Basin (101300N and 641400W) between November 1995 and January 2017. All samples were collected using a 12 bottle rosette system equipped with a CTD (SeaBird, model SBE 25) (Astor and Varela, 2013). Water samples for chl a and phaeopigment analysis were obtained on a monthly basis from the first cast of each of cruise. Samples were collected at 8 standard depths in the upper 100 m (1, 7, 15, 25, 35, 55, 75, and 100 m) following Varela (2013). For phytoplankton species determination by microscopy, 500 ml of seawater were collected at the same depths into HDPE bottles and fixed with a 4\% formalin solution neutralized with sodium tetraborate. Samples were analyzed at the Universidad de Oriente, Venezuela, using the Utermöhl technique (Hasle, 1978) with 100 ml sedimentation chambers and a settling period of 48 h. Identifications and enumerations by microscopy were limited to microphytoplankton (cells $>$20 mm) (Mutshinda et al., 2013).

    % The below is from Taylor et al. 2012 again
    % I can use as template to introduce ERA5, MEIv2 etc.
    To gauge ITCZ changes relevant to Cariaco Basin data, we developed ITCZ indices over South America (20°S to 5°N and 290°W to 320°W) and the Tropical Atlantic (15°S to 20°N and 300°W to 357.5°W) as area-weighted precipitation and convergence, using the precipitation intensity and negative divergence of horizontal winds, respectively. We calculate ITCZ indices using the previously described center of action (COA) approach (39). The NCEP Reanalysis (40) was the source of data for NAO (www.cgd.ucar.edu/cas/ jhurrell/ indices.html), Azores High, and ITCZ analyses. Precipitation intensities were obtained from Global Precipitation Climatology Project (41). In all cases except the NAO, derived data were obtained from the National Oceanic and Atmospheric Administration/Office of Oceanic and Atmospheric Research/Earth System Research Laboratory Physical Sciences Division (NOAA/ OAR/ESRL PSD) (www.esrl.noaa.gov/psd).

    \subsubsection{Climate data}
    %from Taylor et al. 2012
    Zonal wind stress (−u, easterly Trade Winds) across the southern Caribbean Sea drives near-shore upwelling of nutrient-rich waters by Ekman transport. We extracted the -u wind component at 10 m from the ECMWF Reanalysis v5 (ERA5) climate model and data reanalysis at the grid cell closest to the Cariaco Time Series sampling coordinates (10.5°N,64.8°W ?double check?), obtained from \url{https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels} (accessed June 2024)

    The monthly AMO index data (10 year lowpass), defined as an area average of detrended low-pass filtered North Atlantic SST anomalies, were obtained from \url{https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo} (accessed in June, 2024).

    The bi-monthly multivariate El Niño/Southern Oscillation (ENSO) indices (MEI v.2), defined as the time series of the leading combined Empirical Orthogonal Function (EOF) of five different variables (sea level pressure (SLP), sea surface temperature (SST), zonal and meridional components of the surface wind, and outgoing longwave radiation (OLR)) over the tropical Pacific basin (30°S-30°N and 100°E-70°W) were obtained from \url{https://psl.noaa.gov/enso/mei/} (accessed in June, 2024). In our analysis, bi-monthly values are aligned with the tail month (December-January is treated as the value for January).

    \subsubsection{Data processing}
    % Niskin, CTD and Phytoplankton data processing
    - quality control, removed outliers, interpolation, averaging
    All done in R

    \subsection{Data analysis}

    \subsubsection{Z-Scores}
    % from Fraker et al. 2022 - Lake Erie paper
    To visualize how ecosystem components changed through time and to identify similarities in temporal trends among variables, each standardized data point (i.e., annual value) was color-coded based on whether its standardized score was greater or less than the long-term mean (negative anomaly: red; positive anomaly: blue; deeper shades indicating greater values). The variables were then ordered based on their loading onto PCA axis 1 (R package ‘ade4’; Dray and Dufour, 2007; R version 3.6.3; RCoreTeam, 2020) to indicate their degree of change over the time interval and their similarity with one another. Such visualizations have been shown to provide a useful portrayal of ecosystem change (e.g., Choi et al., 2005; Vasilakopoulos et al., 2017; Harvey et al., 2020).

    \subsubsection{BDEF and Diversity}
    % from Ptacnik et al. 2008
    Diversity was defined as the number of genera present in a given sample. Reducing the taxonomic resolution to genus level made the dataset more homogeneous and more robust to differences in skill and effort among the individual phytoplankton analysts.

    % Regression analysis of RUE
    To analyze the variability of phytoplankton diversity across temporal and depth ranges, we interpolated the discrete depth counts and divided the resulting depth profile by 4 equal depth intervals (0-25, 25-50, 50-75 and 75-100 m).

    RUE: We excluded observations with TP concentrations 2 gliter1 (n  15) because they are close to the detection limit, resulting in very imprecise measurements.
    Phytoplankton biomass divided by the limiting resource (total phosphorus) gave the yield or resource use efficiency (RUE) for the community. We estimated RUE both in terms of chlorophyll-a (RUEchl = chlorophyll-a:TP) and carbon biomass (RUEcarb  PPC:TP)

    \subsubsection{Clustering and NMDS}
    % from Winder and Hunter 2008
    Non-metric multidimensional scaling (NMDS) ordination and hierarchical clustering analysis were used to extract gradients in phytoplankton community composition over time based on the Sørenson (Bray–Curtis) distance matrix. NMDS is a powerful tool for ecological community data that include a large portion of zero values because it is not biased by the ‘‘joint absences’’ effect (McCune et al. 2002) and outperformed linear models (such as principal components analysis or redundancy analysis). NMDS is an iterative search for the positions of species and time on few dimensions (axes) that minimizes the departure from monotonicity in the association between distance (dissimilarity) in the original data and ordination space. It plots dissimilar objects far apart in the ordination space and similar objects close to one another. The temporal dynamics of the phytoplankton community was characterized based on biovolume averages to reflect the true contribution of individual species to phytoplankton community composition (Interlandi and Kilham 2001). Temporal change of phytoplankton genera is conducted on a matrix of annual averages (year) by genera biovolume to extract years with similar phytoplankton structure. Data were transformed by square root transformation and double standardized (species by maximum species biovolume and dates to unit totals). NMDS is started from random starting configurations and the solution with the lowest ‘‘stress’’ was selected from 500 runs. Stress measures departure of fit between the ordination and the data and is calculated as a departure from monotonicity between the two series of distances. A stable three-dimensional solution was identified for all ordinations. Orthogonal rotation of the NMDS solution was used to maximize correlations of axes so that the variance of points is maximized on first dimension. Environmental variables were related to the strongest gradients (axes) in species composition and, using overlays, in the ordination space towards which the environmental vectors change most rapidly and to which they have maximal correlations with the ordination configuration. Linear regression was used to relate the environmental data of nutrient concentration, optical and physical (mixing) properties, as well as zooplankton (crustacean) density to the NMDS axis scores (Jovan and McCune 2005). Multivariate analysis was conducted using the vegan package in R.
    % could add indicator species per cluster, as they have done

    \subsubsection{Gradient Forest}
    % from Fraker et al. 2022 - Lake Erie paper
    gradient forest analysis was used to identify and compare among the drivers that predicted the temporal trends within each of the indicator datasets and time intervals examined. Gradient forest analysis is a community-level extension of random forest analysis that aggregates regression trees to quantify turnover in indicator variables across driver (or pressure) variables (Pitcher et al., 2012; Large et al., 2015; Tam et al., 2017). It partitions a goodness-of-fit statistic among each driver variable for each indicator variable, then averages over indicators to produce an overall importance for that driver (R2P). The split values of the regression trees also identify thresholds for the driver variables at which the chosen indicator variables exhibit large changes in value (e.g., community turnover in the fish, bird, and zooplankton abundance indicator datasets). Plots of the cumulative predictive importance of each driver variable can suggest target values for management of key drivers of ecosystem condition (i.e., where thresholds occur; Tam et al., 2017). These analyses were conducted for abundances of each taxa and the fish-derived ecosystem indicators within each time interval (the long, intermediate, and short-term intervals, as well as each 20-year interval), except for zooplankton during the 1969–2018, 1969–1988, and 1979–1998 intervals due to lack of monitoring data during these times. These analyses were conducted using R package ‘gradientforest’ with 500 trees, a maximum of three splits, and correlation threshold of 0.5; \cite{ellis_gradient_2012}.

@@

- Phytoplankton.csv - https://www.bco-dmo.org/dataset/3095
	last accessed Jun 2024 (2.6.2024)
	Troccoli, L., Diaz-Ramos, R., Subero-Pino, S., Muller-Karger, F., Astor, Y., Varela, R., Rueda-Roa, D., Klein, E. (2019) Time-Series of Phytoplankton Taxonomy and Density collected by the CARIACO Ocean Time-Series Project from November 1995 to January 2017. Biological and Chemical Oceanography Data Management Office (BCO-DMO). (Version 1) Version Date 2019-08-14. doi:10.1575/1912/bco-dmo.3095.1 [last accessed June 2024]

- Niskin.csv - https://www.bco-dmo.org/dataset/3093
	last accessed Jun 2024 (2.6.2024)
	Muller-Karger, F., Astor, Y., Scranton, M., Taylor, G., Thunell, R., Varela, R., Benitez-Nelson, C., Buck, K., Fanning, K., Capelo, J., Gutierrez, J., Guzman, L., Lorenzoni, L., Montes, E., Rojas, J., Rondon, A., Rueda-Roa, D., Tappa, E. (2019) Time-series Niskin-bottle sample data from R/V Hermano Gines cruises in the Cariaco Basin from 1995 through 2017 (CARIACO Ocean Time-Series Program). Biological and Chemical Oceanography Data Management Office (BCO-DMO). (Version 1) Version Date 2019-06-06. doi:10.1575/1912/bco-dmo.3093.1 [last accessed June 2024]

- CTD - https://www.bco-dmo.org/dataset/3092
	Muller-Karger, F., Astor, Y., Benitez-Nelson, C., Scranton, M., Taylor, G., Thunell, R., Varela, R., Capelo, J., Guzman, L., Lorenzoni, L., Montes, E., Rojas, J., Rueda-Roa, D. (2019) Time series composite CTD profiles from R/V Hermano Ginés cruises in the Cariaco Basin from 1995 through 2017 (CARIACO Ocean Time-Series Program). Biological and Chemical Oceanography Data Management Office (BCO-DMO). (Version 1) Version Date 2019-06-04 [if applicable, indicate subset used]. doi:10.1575/1912/bco-dmo.3092.1 [access date]

- ERA5 - https://cds.climate.copernicus.eu/cdsapp#!/dataset/10.24381/cds.f17050d7
	Hersbach, H., Bell, B., Berrisford, P., Biavati, G., Horányi, A., Muñoz Sabater, J., Nicolas, J., Peubey, C., Radu, R., Rozum, I., Schepers, D., Simmons, A., Soci, C., Dee, D., Thépaut, J-N. (2023): ERA5 monthly averaged data on single levels from 1940 to present. Copernicus Climate Change Service (C3S) Climate Data Store (CDS), DOI: 10.24381/cds.f17050d7 (Accessed on 02-06-2024)

- AMO - https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo 
	Trenberth, Kevin &, Zhang, Rong & National Center for Atmospheric Research Staff (Eds). Last modified 2023-09-01 "The Climate Data Guide: Atlantic Multi-decadal Oscillation (AMO) and Atlantic Multidecadal Variability (AMV).” Retrieved from [https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo](https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo) on 2024-06-06.
- MEI v.2 - https://psl.noaa.gov/enso/mei/

@@

\begin{figure}
\noindent\includegraphics[width=\textwidth]{fig/BDEF_depth_PLOT.pdf}
\caption{Linear regression of diversity indices to log-transformed Chlorophyll Biomass, Primary Production, Resource Use Efficiency for nitrate ($RUE_N$) and for phosphate ($RUE_P$). $RUE$ is calculated as the ratio of Chlorophyll to the available nutrient ($NO_3$ and $PO_4$ respectively). All variables show a significant increasing linear trend with Genus Richness, but none for the Shannon Index or Pielou\'s Evenness Index.}
\end{figure}

@@

\section{Discussion}
This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem, that is driven by local environmental variability, which is in turn affected by large scale climatic shifts. In the Cariaco basin, long-term climatic oscillations have affected the upwelling regime and ...
In the Cariaco basin, wind-driven seasonal upwelling dictates the nutrient availability in the upper mixed layer, which in turn affects the phytoplankton dynamics \cite{ptacnik_diversity_2008, bopp_response_2005}. In particular, Diatoms are responsive to changes in the mixing regime \cite{huisman_reduced_2006}.
Alterations in physical mixing affected phytoplankton dynamics,
% MEI index relevant (from 2019 REVIEW)
The Caribbean experiences the most marked of the El Niño–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Niño–Southern Oscillation (Enfield & Mayer 1997).

% REALLY ONLY LOOK AT BOTTOM UP HERE; No Top Down effects quantified

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% REFERENCES and BIBLIOGRAPHY
%
%\bibliography{<name of your .bib file>} don't specify the file extension
% don't specify bibliographystyle
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
